---
title: "GO-Enrichment Analysis"
author: "Jonathan Speh"
format:
  html:
    code-fold: true
    code-tools: true
    df-print: paged 
    tbl-cap-location: top
    code-overflow: wrap
    self-contained: true
engine: knitr
execute: 
  cache: true
  output: false
bibliography: references.bib
---

# Libraries

```{r}
#| label: setup
#| message: false
#| warning: false

library(data.table)
library(dplyr)
library(tidyr)
library(enrichR)
library(rrvgo)
library(ggplot2)
source("code/helpers.R")

options(scipen = 10000)
```

# GO - Enrichment analysis

```{r}
#| label: prepare-observed
#| message: false

observed_stopgains <-  fread("processed_data/overlap_stopgain_unique_parsed.csv")
observed_stopgains_phred40 <- observed_stopgains |> 
  filter(PHRED >= 40, 
         AC > 1) # TODO: try without removing singletons?

genelist_observed <- unique(observed_stopgains_phred40$GeneName)

dbs <- c("GO_Biological_Process_2023", "GO_Cellular_Component_2023", "GO_Molecular_Function_2023")

enriched_observed <- enrichr(unique(genelist_observed), dbs)

observed_BP <- process_enrichR(enriched_observed$GO_Biological_Process_2023)
observed_MF <- process_enrichR(enriched_observed$GO_Molecular_Function_2023)
observed_CC <- process_enrichR(enriched_observed$GO_Cellular_Component_2023)

observed_sim_mat_BP <- calculateSimMatrix(observed_BP$GO_ID,
                   orgdb="org.Hs.eg.db",
                   ont = "BP",
                   method = "Rel")

observed_sim_mat_MF <- calculateSimMatrix(observed_MF$GO_ID,
                   orgdb="org.Hs.eg.db",
                   ont = "MF",
                   method = "Rel")

observed_sim_mat_CC <- calculateSimMatrix(observed_CC$GO_ID,
                   orgdb="org.Hs.eg.db",
                   ont = "CC",
                   method = "Rel")


observed_scores_BP <- setNames(-log10(observed_BP$Adjusted.P.value), observed_BP$GO_ID)
observed_scores_MF <- setNames(-log10(observed_MF$Adjusted.P.value), observed_MF$GO_ID)
observed_scores_CC <- setNames(-log10(observed_CC$Adjusted.P.value), observed_CC$GO_ID)



observed_sim_mat_BP_reduced <- reduceSimMatrix(observed_sim_mat_BP,
                                      scores = observed_scores_BP,
                                      threshold = 0.7,
                                      orgdb="org.Hs.eg.db")
observed_sim_mat_MF_reduced <- reduceSimMatrix(observed_sim_mat_MF,
                                      scores = observed_scores_MF,
                                      threshold = 0.7,
                                      orgdb="org.Hs.eg.db")
observed_sim_mat_CC_reduced <- reduceSimMatrix(observed_sim_mat_CC,
                                      scores = observed_scores_CC,
                                      threshold = 0.7,
                                      orgdb="org.Hs.eg.db")



```

```{r}
#| label: prepare-simulated
#| message: false

all_stopgains <-  fread("processed_data/all_unique_cadd_snvs.csv")
simulated_stopgains <- anti_join(all_stopgains[,c("Ident", "relcDNApos", "PHRED", "GeneName")],
                                observed_stopgains[,c("Ident", "relcDNApos", "PHRED",  "GeneName")],
                                by = join_by(Ident, relcDNApos, PHRED))

simulated_stopgains_phred40 <- simulated_stopgains |> filter(PHRED >= 40)

genelist_simulated <- unique(simulated_stopgains_phred40$GeneName)

dbs <- c("GO_Biological_Process_2023", "GO_Cellular_Component_2023", "GO_Molecular_Function_2023")

enriched_simulated <- enrichr(unique(genelist_simulated), dbs)

simulated_BP <- process_enrichR(enriched_simulated$GO_Biological_Process_2023)
simulated_MF <- process_enrichR(enriched_simulated$GO_Molecular_Function_2023)
simulated_CC <- process_enrichR(enriched_simulated$GO_Cellular_Component_2023)

simulated_sim_mat_BP <- calculateSimMatrix(simulated_BP$GO_ID,
                   orgdb="org.Hs.eg.db",
                   ont = "BP",
                   method = "Rel")

simulated_sim_mat_MF <- calculateSimMatrix(simulated_MF$GO_ID,
                   orgdb="org.Hs.eg.db",
                   ont = "MF",
                   method = "Rel")

simulated_sim_mat_CC <- calculateSimMatrix(simulated_CC$GO_ID,
                   orgdb="org.Hs.eg.db",
                   ont = "CC",
                   method = "Rel")


simulated_scores_BP <- setNames(-log10(simulated_BP$Adjusted.P.value), simulated_BP$GO_ID)
simulated_scores_MF <- setNames(-log10(simulated_MF$Adjusted.P.value), simulated_MF$GO_ID)
simulated_scores_CC <- setNames(-log10(simulated_CC$Adjusted.P.value), simulated_CC$GO_ID)



simulated_sim_mat_BP_reduced <- reduceSimMatrix(simulated_sim_mat_BP,
                                      scores = simulated_scores_BP,
                                      threshold = 0.7,
                                      orgdb="org.Hs.eg.db")
simulated_sim_mat_MF_reduced <- reduceSimMatrix(simulated_sim_mat_MF,
                                      scores = simulated_scores_MF,
                                      threshold = 0.7,
                                      orgdb="org.Hs.eg.db")
simulated_sim_mat_CC_reduced <- reduceSimMatrix(simulated_sim_mat_CC,
                                      scores = simulated_scores_CC,
                                      threshold = 0.7,
                                      orgdb="org.Hs.eg.db")


```

```{r}
#| label: plot-ea-bp
#| output: true
#| message: false
#| layout-ncol: 2
#| fig-cap-location: top
#| fig-cap: "Biological Process"


treemapPlot(observed_sim_mat_BP_reduced, title = "Observed")
treemapPlot(simulated_sim_mat_BP_reduced, title = "Simulated")
```

-   The gene list from the simulated SNVs (with phred \> 40) that where not observed was enriched for more regulatory functions, including regulation of cell cycle and protein biosynthesis
-   The gene list form the observed SNVs (with phred \> 40) is enriched for metabolic processes, cation transport and tissue and cell morphology

```{r}
#| label: plot-ea-cc
#| output: true
#| message: false
#| layout-ncol: 2
#| fig-cap-location: top
#| fig-cap: "Cellular Component"

treemapPlot(observed_sim_mat_CC_reduced, title = "Observed")
treemapPlot(simulated_sim_mat_CC_reduced, title = "Simulated")
```

```{r}
#| label: plot-ea-mf
#| output: true
#| message: false
#| layout-ncol: 2
#| fig-cap-location: top
#| fig-cap: "Molecular Function"

treemapPlot(observed_sim_mat_MF_reduced, title = "Observed")
treemapPlot(simulated_sim_mat_MF_reduced, title = "Simulated")
```

# Phenotype enrichment analysis

-   EA of genelists from simulated stopgains, singletons, and stopgains that where observerd more than once using the Mammalian Phenotype Ontology [@smith2012]
-   MP is however based on mouse data ([mouse genome informatics database](https://www.informatics.jax.org/))

```{r}
#| label: phenotype-enrichment

dbs <- c("MGI_Mammalian_Phenotype_Level_4_2021")


genelist_singleton <- observed_stopgains |> filter(PHRED >= 40, 
                                                        AC == 1)

genelist_singleton <- unique(genelist_singleton$GeneName)


genelist_non_singleton <- observed_stopgains |> filter(PHRED >= 40, 
                                                        AC > 1)
genelist_non_singleton <- unique(genelist_non_singleton$GeneName)


singleton_enriched<- enrichr(genelist_singleton, dbs)
non_singleton_enriched <- enrichr(genelist_non_singleton, dbs)
simulated_enriched <- enrichr(genelist_simulated, dbs)
```

```{r}
#| label: proc-pte
#| tbl-cap: 
#|    -   "Phenotype enrichment of genelist from Simulated SNVs"
#|    -   "Phenotype enrichment of genelist from Singletons"
#|    -   "Phenotype enrichment of genelist from non-singleton variants"

singleton_enriched_sig <- singleton_enriched$MGI_Mammalian_Phenotype_Level_4_2021|>
  filter(Adjusted.P.value <= 0.05)

non_singleton_enriched_sig <- non_singleton_enriched$MGI_Mammalian_Phenotype_Level_4_2021|>
  filter(Adjusted.P.value <= 0.05) 

simulated_enriched_sig <- simulated_enriched$MGI_Mammalian_Phenotype_Level_4_2021|>
  filter(Adjusted.P.value <= 0.05) 


simulated_enriched_sig |> arrange(Adjusted.P.value)
singleton_enriched_sig |> arrange(Adjusted.P.value)
non_singleton_enriched_sig |> arrange(Adjusted.P.value)
```

```{r}
#| label: plot-pte-facet
#| output: true

n = 10

pt_enrichment <- rbind(slice_min(simulated_enriched_sig, Adjusted.P.value, n = n, with_ties = FALSE),
      slice_min(singleton_enriched_sig, Adjusted.P.value, n = n, with_ties = FALSE),
      slice_min(non_singleton_enriched_sig, Adjusted.P.value, n = n, with_ties = FALSE)) |>
  mutate(flag = factor(rep(c("simulated", "singleton", "non-singleton"), each = n),
                       levels = c("simulated", "singleton", "non-singleton"))) |>
  rowwise() |>
  mutate(Overlap = eval(parse(text = Overlap))) |>
  separate_wider_delim(Term, "MP:", names = c("Term", "ID"))
  


# pt_enrichment |>
#   ggplot(aes(x = reorder(Term, Overlap), y = Overlap, fill = Adjusted.P.value)) +
#   geom_bar(stat = "identity") +
#   coord_flip() +
#   facet_wrap(~flag) + 
#   scale_fill_gradient(low = "blue", high = "red") +
#   scale_y_continuous(labels = c(0, 0.25 , 0.5, 0.75, 1)) +
#   theme(axis.title.y = element_blank())
  
pt_enrichment |>
  mutate(log10_P = log10(Adjusted.P.value)) |>
  ggplot(aes(x = reorder(Term, -log10_P), y = log10_P, fill = Overlap)) +
  geom_bar(stat = "identity") +
  coord_flip() +
  facet_wrap(~flag, scales = "free_x") + 
  scale_fill_gradient(low = "blue", high = "red") +
  scale_y_reverse() +
  theme(axis.title.y = element_blank(),
        axis.title.x = element_text("log10 P-Value" ))



```

-   Gene lists from simulated SNVs and Singletons are enriched for more "lethal" phenotypes than the gene list from non-singleton SNVs

```{r}
#| label: plot-pte-individual
#| eval: false
#| fig-cap: 
#|    -   "Simulated"
#|    -   "Singletons"
#|    -   "Non-Singletons"
#| fig-cap-location: top


simulated_enriched_sig |> arrange(Adjusted.P.value) |>
  head(n = 20) |>
  rowwise() |>
  mutate(Overlap = eval(parse(text = Overlap))) |>
  separate_wider_delim(Term, "MP:", names = c("Term", "ID")) |>
  ggplot(aes(x = reorder(Term, Adjusted.P.value), y = Overlap, fill = Adjusted.P.value)) +
  geom_bar(stat = "identity") +
  coord_flip() 

singleton_enriched_sig |> arrange(Adjusted.P.value) |>
  head(n = 20) |>
  rowwise() |>
  mutate(Overlap = eval(parse(text = Overlap))) |>
  separate_wider_delim(Term, "MP:", names = c("Term", "ID")) |>
  ggplot(aes(x = reorder(Term, Adjusted.P.value), y = Overlap, fill = Adjusted.P.value)) +
  geom_bar(stat = "identity") +
  coord_flip() 

non_singleton_enriched_sig |> arrange(Adjusted.P.value) |>
  head(n = 20) |>
  rowwise() |>
  mutate(Overlap = eval(parse(text = Overlap))) |>
  separate_wider_delim(Term, "MP:", names = c("Term", "ID")) |>
  ggplot(aes(x = reorder(Term, Adjusted.P.value), y = Overlap, fill = Adjusted.P.value)) +
  geom_bar(stat = "identity") +
  coord_flip() 


```
